---
title: "APA Homework #1"
author: "J. Adesola, L. Jia, M. Xu, Yiyan Zhang, C. Zu"
date: "10/23/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Load library
library(ggplot2)
```

## I. Data Overview
### The bikeshare dataset has 731 records that contains users bike rental activities from Jan 2011 to Dec 2012. There are 7 categorical variables and 4 continuous variables that have potential influences on the number of rental bike users. We are going to build a linear regression model to explore elements that affect the ridership. 
```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Load data
df <- read.csv("HW1 bikeshare.csv")
#str(df)

# Convert dteday as date variable
df$dteday <- as.Date(df$dteday)
```

## II. Independent and Dependent Variables Selection
### We have 3 measurements for the number of users, among which "cnt" is selected as the dependent variable for building the model since it reflects both registered and casual users' rental behaviour. The graph below shows the daily volume of bike riders over the two-year time period, the red line is the total average.
```{r, fig.height = 8, fig.width = 20, echo = FALSE, warning = FALSE, message = FALSE}
# Plot time series for number for users
ggplot(df, aes(x = dteday, y = cnt)) +
  geom_line() +
  geom_hline(aes(yintercept = mean(cnt)), color = "red", linetype = "dashed", size = 1) +
  labs(title = "Time Series for Number of Users", x = "Date", y = "Count")
```

### We found that "season" and "mnth" have a high correlation of 0.83 (a high correlation means two variables have a strong relationship and their impacts in the model will be similar). Between these two variables, we selected "season" to be included in building the regression model since "season" has fewer levels and is a more representative variable when making decisions related to time dimension. "temp" and "atemp" have a high correlation of 0.99. We chose "atemp" because it measures the users' feelings of temperature. We chose "holiday" as opposed to "workingday" because they are interchangeable variables and either could be used.
```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Correlation matrix for variables
# round(cor(df[,3:11]),2)
```


## III. Data Cleaning
### Outliers (abnomal observations) are detected in "windspeed" and "hum" variables, and we replace them with the respective mean values.
```{r, echo = FALSE, warning = FALSE, message = FALSE}
# # Plot boxplot for numeric variables
# boxplot(df$temp,  main = "Boxplot for temp", xlab = "temp", ylab = "Normalized value")
# boxplot(df$atemp,  main = "Boxplot for atemp", xlab = "atemp", ylab = "Normalized value")
# boxplot(df$hum,  main = "Boxplot for hum", xlab = "hum", ylab = "Normalized value")
# boxplot(df$windspeed,  main = "Boxplot for windspeed", xlab = "windspeed", ylab = "Normalized value")
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Mean imputation for outliers
num <- c("hum", "windspeed")

df[num] <- data.frame(lapply(
  df[num], function(x) {
    ifelse(x < (mean(x, na.rm = TRUE) - 3 * sd(x, na.rm = TRUE)) | 
           x > (mean(x, na.rm = TRUE) + 3 * sd(x, na.rm = TRUE)), 
           NA, x)
    }))

df[num] <- data.frame(lapply(df[num], function(x) {
  ifelse(is.na(x), mean(x, na.rm = TRUE), x)
  }))
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Convert categorical variables to factors
df$season <- as.factor(df$season)
df$yr <- as.factor(df$yr)
df$mnth <- as.factor(df$mnth)
df$holiday <- as.factor(df$holiday)
df$weekday <- as.factor(df$weekday)
df$workingday <- as.factor(df$workingday)
df$weathersit <- as.factor(df$weathersit)
```

## IV. Frequency and Distribution of Independent Variables
```{r, fig.height = 4, fig.width = 15, echo = FALSE, warning = FALSE, message = FALSE}
# Plot histograms for numerical variables
par(mfrow = c(1,3))
hist(df$atemp)
hist(df$windspeed)
hist(df$hum)
```

```{r, fig.height = 10, fig.width = 15, echo = FALSE, warning = FALSE, message = FALSE}
# Plot bar graphs for categorical variables
par(mfrow = c(3, 3))

c1 <- table(df$season)
barplot(c1, xlab = "Season")

c2 <- table(df$yr)
barplot(c2, xlab = "Year")

c3 <- table(df$mnth)
barplot(c3, xlab = "Month")

c4 <- table(df$holiday)
barplot(c4, xlab = "Holiday")

c5 <- table(df$weekday)
barplot(c5, xlab = "Weekday")

c6 <- table(df$workingday)
barplot(c6, xlab = "Working Day")

c7 <- table(df$weathersit)
barplot(c7, xlab = "Weather")
```

## V. Regression Model
```{r, echo = FALSE, warning = FALSE, message = FALSE}
# reg_cnt = lm(cnt ~ hum + atemp + windspeed + season + holiday + weathersit + weekday, df)
# summary(reg_cnt)
```
### Input variables - hum, atemp, windspeed, season, holiday, weathersit, weekday
### Output variables - cnt(count of total rental bikes including both casual and registered)
### The final regression model is: 
### cnt = 2817.6 + -2905.3 * hum + 6707.2 * atemp + -3034.1 * windspeed + 955.9 * season2 + 606.2 * season3 + 1499.4 * season4 + -589.8 * holiday1 + -193.1 * weathersit2 + -1777.8 * weathersit3 + 187.4 * weekday1 + 271.3 * weekday2 + 365.9 * weekday3 + 337.6 * weekday4 + 392.2 * weekday5 + 413.9 * weekday6
### According to the regression summary output, there is strong evidence showing our model is statistically significant and can be used towards predicting rental bike users' behaviours. 
### Humidity, windspeed and bad weather (e.g. misty, cloudy, rainy and snowy) have a negative impact on the total volume of rental bike users. A unit increase in atemperature will lead to an increase of around 134 (6706.2/50=134) rental bike users.  
### Among the four seasons, winter has the most impact on bike rental activities. Winter can bring up the volume of rental bike users by 1499.4 whereas fall only leads to an increase of 606.2, which is half of the impact from winter's.
### As for weekdays, Friday and Saturday have a relative high influence on the ridership. 
### The intercept of our model (2817.6) means if we ignore all the influencing factors, there should be 2817 bike rentals per day.


## VI. Insights and Recommendations
### Our finding from the regression model indicates that there is a positive relationship between bike usage and feeling temperature for rental bike users. However, winter has the strongest positive impact on ridership. Spring and fall do not affect bike rentals as much. 
### People tend to rent bikes during working day as opposed to holidays. From our regression model we can assume people prefer to rent bikes when they need to commute to work. 

### Recommendations: 
### 1. Increase the supply of rental bikes during the winter and on working days to meet the corresponding increase in demand and maximize revenue. 
### 2. Raise marketing efforts during the spring and fall seasons to attract more bike users. Sustainability is a hot topic in recent years. The company could incorporate the concept of sustainable transportation in marketing campaign to raise people's awareness in bike rental activities during the short-demand seasons.
### 3. Potential partnerships could be formed with lesuire destinations (e.g. theme parks). Since there is less ridership during holidays, we could offer holiday rental discounts or promotion bundles (e.g. discount tickets if you ride to the theme park) to encourge people to choose rental bikes as their primary mode of transportation, thus stimulating the revenue during holidays.